Clean GTEx Genotypes for TWAS Expression Imputation

### 2019-12-11 code/GTExPreProcess.Rmd
GTEx Processing

```{r libs, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  results='hide')
#library(reticulate)
```
##Before running 
Run the following in docker exec -it <CONTAINER> /bin/bash 
chmod 777 /root/
mv TWAS/ /home/<USR>/TWAS/
chown -hR <GID>:<USR> /home/<USR>/TWAS/
Create a ~/.synapseConfig file in the container, but do not push a container containing your credentials file to Docker Hub!!!

@@

~/TWAS/bin/plink --threads 10 --bfile All_CEU_ToTrainForTWAS --write-snplist

@@

    ~/TWAS/bin/plink --threads 10 --file chr$i --extract plink.snplist -make-bed --out chr$i\_filt
    rm chr$i.ped
    rm chr$i.map
    rm chr$i.log

@@

~/TWAS/bin/plink --threads 10 --bfile chr1_filt --merge-list merge.txt --make-bed --out GTEx_FiltSNP_Data
rm merge.txt
rm Log.txt 
rm chr*_filt.*

@@

@@

CODE <- syn_temp$store(synapseclient$Folder(name = "GTExGenotypes", parentId = parentId))

@@

@@

```{bash CleanClean, results='hide', echo=TRUE, cache=TRUE}
rm *.bed
rm *.fam
rm *.bim
rm *.nosex
rm *.log
rm *.snplist
```

>>>>

```{r libs, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  results='hide')
library(reticulate)
#BiocManager::install("Rgraphviz")
#BiocManager::install("graph")
#install.packages('CePa')
#BiocManager::install("phantasus")
```
##Before running 
Run the following in docker exec -it <CONTAINER> /bin/bash 
chmod 777 /root/
mv TWAS/ /home/<USR>/TWAS/
chown -hR <GID>:<USR> /home/<USR>/TWAS/
Create a ~/.synapseConfig file in the container, but do not push a container containing your credentials file to Docker Hub!!!

@@

~/TWAS/bin/plink --threads 12 --bfile All_CEU_ToTrainForTWAS --write-snplist

@@

    ~/TWAS/bin/plink --threads 12 --file chr$i --extract plink.snplist -make-bed --out chr$i\_filt
    rm chr$i.ped
    rm chr$i.map
    rm chr$i.log

@@

~/TWAS/bin/plink --threads 12 --bfile chr1_filt --merge-list merge.txt --make-bed --out GTEx_FiltSNP_Data
rm merge.txt
rm Log.txt 
rm chr*_filt.*

@@

## Pull GTEx WGS Data From Synapse and Prep for Imputation
This will need a bunch of space!
```{bash Pull_WGS, results='hide', echo=TRUE, cache=TRUE}
synapse get syn21417881
gunzip GTEx_Analysis_2016-01-15_v7_WholeGenomeSeq_635Ind_PASS_AB02_GQ20_HETX_MISS15_PLINKQC.vcf.gz

synapse get syn20835007

awk '{ print $1"\t"$4 }' All_CEU_ToTrainForTWAS.bim > Positions.txt

vcftools --gzvcf GTEx_Analysis_2016-01-15_v7_WholeGenomeSeq_635Ind_PASS_AB02_GQ20_HETX_MISS15_PLINKQC.vcf.gz --positions Positions.txt --plink --out Init_Filtered

~/TWAS/bin/plink --threads 12 --file Init_Filtered --make-bed --out temp

### ReName the MSBB SNPS so the can be combined
```{python ReNamer_ONE, echo=F, results='hide'}
import sys
import os
import re

InFile = 'All_CEU_ToTrainForTWAS.bim'
InFileb = 'temp.bim' #LDREF_FiltSNPs.bim

RS_IDs = {}
RS_Pos = {}

F1 = open(InFile)
for line in F1:
	LINE = line.rstrip('\r\n')
	lst = LINE.split('\t')
	RS_IDs[ ''.join([lst[0], '_', lst[3], ':', lst[4], '-', lst[5]] ) ] = lst[1]
	RS_IDs[ ''.join([lst[0], '_', lst[3], ':', lst[5], '-', lst[4]] ) ] = lst[1]
	#RS_IDs[ ''.join([lst[0], '_', lst[3]] ) ] = lst[1]
	#RS_Pos[ ''.join([lst[0], '_', lst[3], ':', lst[4], '-', lst[5]] ) ] = lst[2]
F1.close()

#Open Output
OUT = open( ''.join([ InFileb.replace(".bim", ""), 'Renamed.bim']), 'w' )
#OUT2 = open( ''.join([ InFileb.replace(".bim", ""), 'WTF.bim']), 'w' )
F1 = open(InFileb)
for line in F1:
	LINE = line.rstrip('\r\n')
	lst = LINE.split('\t')
	foo = ''.join([lst[0], '_', lst[3], ':', lst[5], '-', lst[4]] )
	#foo = ''.join([lst[0], '_', lst[3] ] ) 
	if foo in RS_IDs:
	  Name = RS_IDs[  ''.join([lst[0], '_', lst[3], ':', lst[4], '-', lst[5] ] ) ]
	  #Name = RS_IDs[  ''.join([lst[0], '_', lst[3] ] ) ]
	  Entry = '\t'.join([ lst[0], Name, lst[2], lst[3], lst[4], lst[5] ])
	  print >>OUT, Entry
	else:
	  print >>OUT, Entry

F1.close()
OUT.close()
#OUT2.close()
```

```{bash FinalizeGenoType, results='hide', echo=TRUE, cache=TRUE}
mv tempRenamed.bim temp.bim

#Remove all Repeated SNPs 9I have no Idea how imputation will handle them
~/TWAS/bin/plink --threads 12 --bfile All_CEU_ToTrainForTWAS --write-snplist
~/TWAS/bin/plink --threads 12 --bfile temp --extract plink.snplist --list-duplicate-vars --make-bed --out Test
~/TWAS/bin/plink --threads 12 --bfile Test --list-duplicate-vars suppress-first --out Duplicates
~/TWAS/bin/plink --threads 12 --bfile Test --exclude Duplicates.dupvar --make-bed --out UniqueGTEx_GWAS_ForImpute

rm Test*
rm *.snplist
rm temp.*
rm Init_Filtered.*
rm Duplicates.*
rm positions.txt

## Pull MetaData & Expression
#_# wget https://storage.googleapis.com/gtex_analysis_v8/multi_tissue_eqtl_data/GTEx_Analysis_v8.metasoft.txt.gz
#_# gunzip GTEx_Analysis_v8.metasoft.txt.gz

wget https://storage.googleapis.com/gtex_analysis_v8/annotations/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt

wget https://storage.googleapis.com/gtex_analysis_v8/rna_seq_data/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz
gunzip GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz
```

```{r GTExPheno, cache=FALSE, echo=FALSE}
#ID Brain INDV
Meta <- as.data.frame( data.table::fread(file='GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt', header=T) )
table(Meta$SMTS)

TissueTypes <- names( table(Meta[ Meta$SMTS == 'Brain', ]$SMTSD) )
TissueTypes <- TissueTypes[ grepl( "Cerebell", TissueTypes ) == F ]

BrainMeta <- Meta[ Meta$SMTSD %in% TissueTypes, ]
IDs <- BrainMeta$SAMPID

GTEX_IDs <- paste0( do.call( rbind, strsplit( IDs,'-' ))[,1], '-', do.call( rbind, strsplit( IDs,'-' ))[,2] )
uGTEX_IDs <-GTEX_IDs[ !duplicated(GTEX_IDs) ]

FAM <- as.data.frame( data.table::fread(file='UniqueGTEx_GWAS_ForImpute.fam', header=F) )
row.names(FAM) <- FAM$V1

BRAIN_INDV <- FAM[ (FAM$V1 %in% uGTEX_IDs) == T, 1:2 ]
write.table( BRAIN_INDV, file = "GTEX_Plink_Indvs.txt", row.names = F, col.names = F, sep='\t', quote=F)

####################################################################################################
#Get Expression
EXP <- phantasus::read.gct('GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct')

exp <- EXP@assayData$exprs
bMeta <- Meta[ Meta$SAMPID %in% IDs, ]
bexp <- exp[ , (colnames(exp) %in% IDs) ==T ] 
row.names( bMeta ) <- bMeta$SAMPID
bMeta <- bMeta[ colnames(bexp), ]

#Filter for relevant genes
row.names(bexp) <- do.call( rbind, strsplit(row.names(bexp), '\\.'))[,1]
Genes <- as.character(as.data.frame(data.table::fread( syn_temp$get('syn21011180')$path, header = T))[,1])
bexp <- bexp[ Genes[ Genes %in% row.names(bexp) ], ]

#Scale the values
ScalerTable <- table( bMeta$SAMPID, bMeta$SMTSD)
ScalerTable <- as.data.frame(ScalerTable)

Scaled_bexp <- as.data.frame( matrix(0,dim(bexp)[1],0) )
row.names( Scaled_bexp ) <- row.names(bexp)

Scaled_bexp_Log2 <- as.data.frame( matrix(0,dim(bexp)[1],0) )
row.names( Scaled_bexp_Log2 ) <- row.names(bexp)

for( Type in names(table(ScalerTable$Var2)) ){
  #print(head( bexp[ ,as.character(ScalerTable[ ScalerTable$Var2 == Type & ScalerTable$Freq == 1,  ]$Var1) ][,1:5]))
  Scal <- t(scale(t(bexp[ ,as.character(ScalerTable[ ScalerTable$Var2 == Type & ScalerTable$Freq == 1,  ]$Var1)] ))) 
  Scaled_bexp <- cbind(Scaled_bexp, Scal)

  Scal_log <- t(scale(t(log2(bexp[ ,as.character(ScalerTable[ ScalerTable$Var2 == Type & ScalerTable$Freq == 1,  ]$Var1) ]+1)))) 
  Scaled_bexp_Log2 <- cbind(Scaled_bexp_Log2, Scal_log)
}

hist(as.numeric(Scaled_bexp['ENSG00000225630',]), breaks =80)
hist(as.numeric(Scaled_bexp_Log2['ENSG00000225630',]), breaks =80)

write.table(Scaled_bexp, file = "Init_GETx_Scaled_Expression.tsv", sep="\t", row.names = T, col.names = T, quote = F)
write.table(Scaled_bexp_Log2, file = "Init_GETx_ScaledLog2Plus1_Expression.tsv", sep="\t", row.names = T, col.names = T, quote = F)

bMeta$PTID <- paste0( do.call( rbind, strsplit( colnames(Scaled_bexp), '-'))[,1], '-', do.call( rbind, strsplit( colnames(Scaled_bexp), '-'))[,2] )
write.table(bMeta, file = "GETx_BrainMeta.tsv", sep="\t", row.names = T, col.names = T, quote = F)

table(FAM$V1 %in% bMeta$PTID)

@@

CODE <- syn_temp$store(synapseclient$Folder(name = "GTExGenotypes", parentId = parentId))
CODEalt <- syn_temp$store(synapseclient$Folder(name = "GTExPreProcess", parentId = parentId))

@@

# Set annotations
EXP.all.annotations = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay  = 'RNAseq',
  tissueTypeAbrv  = c(gsub( 'Brain - ', "", names(table(bMeta$SMTSD)))), 
  study = 'GTEx', 
  organism = 'HomoSapiens',
  consortium  = 'GTEx',
  normalizationStatus = TRUE,
  normalizationType = 'Scaled TPM',
  rnaquantification = 'UNK',
  genomeAssemblyID = 'GRCh38'
)

Basic.annotations = list(
  dataType = 'MetaData',
  dataSubType = 'GTEx Sample Meta Data',
  study = 'GTEx', 
  organism = 'HomoSapiens',
  consortium  = 'GTEx'
)

@@

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='UniqueGTEx_GWAS_ForImpute.fam', name = 'Filtered GTEx Genotype FAM file from GWAS Data for TWAS Imputation', parentId=CODE$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = Var.all.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='UniqueGTEx_GWAS_ForImpute.bed', name = 'Filtered GTEx Genotype BIM file from GWAS Data for TWAS Imputation', parentId=CODE$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = Var.all.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='UniqueGTEx_GWAS_ForImpute.bim', name = 'Filtered GTEx Genotype BED file from GWAS Data for TWAS Imputation', parentId=CODE$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = Var.all.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='Init_GETx_Scaled_Expression.tsv', name = 'TPM GTEx Brain Expression for all Profiles Z-Scaled within Tissue', parentId=CODEalt$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = EXP.all.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='Init_GETx_ScaledLog2Plus1_Expression.tsv', name = 'Log2 TPM+1 GTEx Brain Expression for all Profiles Z-Scaled within Tissue', parentId=CODEalt$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = EXP.all.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='GETx_BrainMeta.tsv', name = 'Brain Sample Meta Data', parentId=CODEalt$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
syn_temp$setAnnotations(ENRICH_OBJ, annotations = Basic.annotations)

ENRICH_OBJ <- syn_temp$store( synapseclient$File( path='GTEX_Plink_Indvs.txt', name = 'Genotype IIDs with a BrainExpression Profile', parentId=CODEalt$properties$id ), used = Syns_Used, activityName = activityName, executed = thisFile, activityDescription = activityDescription)
```

```{bash CleanClean, results='hide', echo=TRUE, cache=TRUE}
rm *.bed
rm *.fam
rm *.bim
rm *.nosex
rm *.log
rm *.snplist
rm *.tsv
rm *.txt
rm *.gct
rm *.gz
```
